Gains and losses in rheumatoid arthritis data from inferCNV

### 2024-02-16 infercnv/reports/gains_and_losses.Rmd
move sce wrangling out of rmd, make gains_and_losses.rmd accept different cell_group_name values

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

```{r load_obj, results = 'asis'}
# dirs
data_dir <- '/lustre/scratch126/casm/team268im/mp34/scRNAseq_data/RA_ZhangEtal2023/processed_output/'

# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

# convert obj to sce
sce <- xfun::cache_rds({

  # load CTAP and metadata
  ctap <-
    readr::read_tsv('data/Zhang2023/CTAP_donor_mapping.tsv') %>%
    dplyr::select(sample = donor, CTAP)
  md <- 
    readr::read_tsv('data/Zhang2023/SYNAPSE_METADATA_MANIFEST.tsv') %>%
    dplyr::distinct()

  # load object
  obj <- 
    readRDS(file.path(data_dir, 'all_cells_reference.rds'))

  # load rna data
  assays <-
    list(
      counts = 
        readRDS(file.path(data_dir, 'raw_mRNA_count_matrix.rds')),
      logcounts = 
        readRDS(file.path(data_dir, 'qc_mRNA_314011cells_log_normalized_matrix.rds')))

  # get cell order from counts
  cell_order <- colnames(assays$logcounts)

  # filter and reorder to cells that pass qc from counts
  assays$counts <- 
    assays$counts[rownames(assays$logcounts), 
                  cell_order]

  # generate coldata
  col_data <-
    obj$meta_data[cell_order, ]

  # populate sce
  sce <- 
    SingleCellExperiment::SingleCellExperiment(
      assays = assays,
      colData = col_data
    )

  # name, reorder, and add umap embeddings
  umap <- obj$umap$embedding
  rownames(umap) <- rownames(obj$meta_data)
  umap <- umap[cell_order, ]
  reducedDim(sce, 'UMAP_uwot') <- umap

@@

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(col_data, cnvs, by = 'row.names', all.x = T, sort = F)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

 # remove NA's for inferCNV analysis
 sce <- sce[, !is.na(sce$proportion_cnv_chr1)]
```

@@

```{r get_hmm}
cnv_regions <- xfun::cache_rds({
  cnv_regions <- 
    list.files(
      params$infercnv_dir,
      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        readr::read_tsv()
      }) %>%
    dplyr::bind_rows(.id = 'sample') %>%
    # refactor chromosomes
    dplyr::mutate(
      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
      direction = dplyr::case_when(
        state < 3 ~ 'loss',
        state > 3 ~ 'dupli',
        TRUE ~ 'neutral'),
      region = gsub('.*region_', '', cnv_name) %>% as.numeric())
}, file = 'cnv_regions', rerun = params$rerun)
```

```{r plot_cnv_regions_pred}
pdf('test.pdf', height = 25)
cnv_regions %>%
  dplyr::filter(direction != 'neutral') %>% 
  ggplot(aes(x = region)) +
  geom_bar(position = 'stack') +
  facet_wrap(~ chr, ncol = 1, scales = 'free') +
  theme_classic() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(legend.position = 'none')
dev.off()
```

@@

First, we plot UMAPs by celltype.

```{r plot_umaps}
dittoSeq::dittoDimPlot(sce, 'cell_type', size = 0.2) +
  coord_fixed()
```

Then, we visualise gains and loss proportions for each chromosome on the UMAP.

```{r plot_chr_gains, fig.width = 10, fig.height = 5}
pdf('test.pdf')
paste0('chr', 1:22) %>% head(2) %>%
  purrr::map(function(chr) {
    sce %>%
      Nebulosa::plot_density(
        paste('proportion', c('dupli', 'loss'), chr, sep = '_'),
        size = 0.5) 
  })
```

@@

```{r plot_gains_and_losses_by_ct}
# at single-cell level
sce$cell_type[!is.na(sce$infercnv)] %>%
  unique() %>%
  purrr::map(function(ct) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[!is.na(sce$proportion_cnv_chr1) & sce$cell_type == ct, ]
        if (nrow(p_dat) > 0) {
          p_mat <-
            p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
          p_annot <- 
            p_dat[, c('sample'), drop = F] %>%
            as.data.frame()
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ifelse(nrow(p_mat) > 65536, F, T),
            legend = F,
            main = paste0('celltype: ', ct, '\ndirection: ', direction)) %>%
            print()
        }
      })
  })
```

@@

```{r plot_gains_and_losses_by_id}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::group_by(sample, cell_type) %>%
  dplyr::filter(!is.na(proportion_cnv_chr1)) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample_x_ct = paste0(sample, '_', cell_type)) %>%
  tibble::column_to_rownames('sample_x_ct')

# split gain and loss
p_dat <-
  c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_type')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction)) %>%
      print()
  })
```

@@

```{r chr7_prop_dist}
# prep data
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::filter(!is.na(proportion_dupli_chr7)) %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(n_cells = dplyr::n())

# define functions
plot_prop_chr7_gain <- function(p_dat, title) {
  plot_func <- function(p_dat, position_arg) {
    p_dat %>%
      ggplot() +
      geom_bar(aes(x = cell_type, 
                   fill = proportion_dupli_chr7, 
                   group = proportion_dupli_chr7),
               position = position_arg) +
      scale_x_discrete(guide = guide_axis(angle = -45)) +
      theme_classic() +
      scale_fill_viridis_b(n.breaks = 9, limits = c(0, 1)) 
  }

@@

We will define cells with chr7 gain as those affecting >90% of genes. We will pseudobulk by patient, celltype, and chr7 ploidy to produce a reduced object.

```{r pseudobulk_chr7_gain}
reduced_sce <- xfun::cache_rds({

  # we assign chr7 status as 
  # >80% of genes affected by a duplication -> chr7+
  # <20% of genes affected by a duplication -> chr7 WT
  sce@colData$chr7_gain <-
    ifelse(sce$proportion_dupli_chr7 > 0.8,
           1, ifelse(sce$proportion_dupli_chr7 < 0.2, 0, NA))

  # pseudobulk the sce object by patient and chr7 ploidy
  # subset to stromal cells
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, !is.na(sce$sample) & 
            !is.na(sce$chr7_gain) &
            !is.na(sce$cell_type)], 
      group_by = vars(sample, cell_type, chr7_gain)
    )

@@

We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status and the patient as factors.

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
  glmGamPoi::glm_gp(
    reduced_sce,
    design = ~ chr7_gain + sample + cell_type,
    verbose = T
  )
  fit
}, file = 'fit.rds', rerun = params$rerun)

@@

  # test differences
  de_res <-
    glmGamPoi::test_de(
      fit, 
      contrast = 
        cond(chr7_gain = 1, cell_type = 'Stromal cells') -
        cond(chr7_gain = 0, cell_type = 'Stromal cells')
    )

@@

```{r plot_de}
# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = log2(1.5), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```

>>>>

---
title: "Gains and losses in rheumatoid arthritis data from inferCNV"
author: "Alexandra Tidd"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    fig_width: 8
    keep_md: true
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    theme: lumen
params:
  sce: NULL
  cell_group_name: NULL
  infercnv_dir: NULL
  cache_dir: NULL
  rerun: true
---

@@

```{r load_sce, results = 'asis'}
# load gene coords
coords <-
  readr::read_tsv('data/hg38_gencode_v27.txt', col_names = c('name', 'chr', 'start', 'stop'))

# load CTAP and metadata
ctap <-
  readr::read_tsv('data/Zhang2023/syn26710600/CTAP_donor_mapping.tsv') %>%
  dplyr::select(sample = donor, CTAP)
md <- 
  readr::read_tsv('data/Zhang2023/syn26710600/SYNAPSE_METADATA_MANIFEST.tsv') %>%
  dplyr::distinct()

# load sce, add inferCNV 
sce <- xfun::cache_rds({

  # load sce
  sce <- readRDS(params$sce)

@@

  # left join cnvs to coldata, reorder, restore rownames
  col_data <- merge(sce@colData, cnvs, by = 'row.names', all.x = T, sort = F)
  rownames(col_data) <- col_data[, 1]
  col_data <- 
    col_data[colnames(sce), -1] %>%
    DataFrame()

@@

# remove NA's for inferCNV analysis
sce <- sce[, !is.na(sce$proportion_cnv_chr1)]

 # assign cell group
sce@colData$cell_group <-
  sce@colData[, params$cell_group_name]
```

@@

```{r get_hmm}
cnv_regions <- xfun::cache_rds({

  # get HMM predictions
  cnv_regions <- 
    list.files(
      params$infercnv_dir,
      pattern = 'HMM_CNV_predictions.HMMi6.leiden.hmm_mode-subclusters.Pnorm_0.5.pred_cnv_regions.dat',
      recursive = T) %>%
    purrr::set_names(., dirname(.)) %>%
    purrr::map(function(file) {
      file.path(params$infercnv_dir, file) %>% 
        readr::read_tsv(show_col_types = F)
      }) %>%
    dplyr::bind_rows(.id = 'sample') %>%
    # refactor chromosomes
    dplyr::mutate(
      chr = chr %>% {forcats::fct_reorder(., gsub('chr', '', .) %>% as.numeric())},
      direction = dplyr::case_when(
        state < 3 ~ 'loss',
        state > 3 ~ 'dupli',
        TRUE ~ 'neutral'),
      ploidy = dplyr::case_when(
        state == 1 ~ '0x',
        state == 2 ~ '0.5x',
        state == 3 ~ '1x',
        state == 4 ~ '1.5x',
        state == 5 ~ '2x',
        state == 6 ~ '>2x',
        TRUE ~ NA
        ) %>% forcats::fct_reorder(state),
      region = gsub('.*region_', '', cnv_name) %>% as.numeric(),
      cell_group = cell_group_name %>% gsub('\\..*', '', .))

  # calculate region midpoints for plotting
  cnv_regions$midpoint <-
    (cnv_regions$start + cnv_regions$end) / 2

  # return
  cnv_regions

}, file = 'cnv_regions.rds', rerun = params$rerun)
```

```{r plot_cnv_regions_pred, fig.height = 20, fig.width = 10}
cnv_regions %>%
  ggplot(aes(x = midpoint, fill = ploidy, group = ploidy)) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'dupli'),
               aes(y = ..density..), 
               position = position_stack(), 
               colour = NA) +
  geom_density(data = cnv_regions %>% dplyr::filter(direction == 'loss'),
               aes(y = -..density..), 
               position = position_stack(reverse = T), 
               colour = NA) +
  ggh4x::facet_grid2(chr ~ cell_group, scales = 'free', independent = 'x') +
  theme_bw() +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c('0x' = '#00008B', '0.5x' = '#b4b0ff', '1' = 'white', 
               '1.5x' = '#ffc2c2', '2x' = '#d64242', '>2x' = '#8B0000')) 
```

@@

First, we plot UMAPs by cell grouping.

```{r plot_umaps}
dittoSeq::dittoDimPlot(sce, params$cell_group_name, size = 0.2) +
  coord_fixed()
```

Then, we visualise gains and loss proportions for each chromosome on the UMAP. The `Nebulosa` package is used to show the values as densities.

```{r plot_chr_gains, fig.width = 10, fig.height = 4, results = 'hide'}
paste0('chr', 1:22) %>% 
  purrr::map(function(chr) {
    events <-
      paste('proportion', c('dupli', 'loss'), chr, sep = '_')
    sce %>%
      Nebulosa::plot_density(events, size = 0.5) %>%
      print()
    reducedDim(sce, 'UMAP_uwot') %>%
      cbind(sce@colData[, events]) %>%
      tibble::as_tibble() %>%
      tidyr::pivot_longer(3:4) %>%
      dplyr::arrange(value) %>%
      ggplot(aes(x = UMAP1, y = UMAP2, colour = value)) + 
      geom_point(size = 0.1) +
      coord_fixed() +
      theme_classic() +
      scale_color_viridis_c() +
      facet_grid(~ name) %>%
      print()
  })
```

@@

```{r plot_gains_and_losses_by_ct, results = 'hide'}
# at single-cell level
sce@colData[, params$cell_group_name] %>%
  unique() %>%
  purrr::map(function(cgn) {
    c('dupli', 'loss') %>%
      purrr::map(function(direction) {
        p_dat <-
          sce@colData[sce@colData[, params$cell_group_name] == cgn, ]
        if (nrow(p_dat) > 0) {
          p_mat <-
            p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
          p_annot <- 
            p_dat[, c('sample'), drop = F] %>%
            as.data.frame()
          pheatmap::pheatmap(
            p_mat, 
            show_rownames = F, 
            annotation_row = p_annot,
            # hclust error: size cannot be NA nor exceed 65536
            cluster_rows = ((nrow(p_mat) < 65536) == T),
            legend = F,
            main = paste0(params$cell_group_name, ': ', cgn, '\ndirection: ', direction)) %>%
            print()
        }
      })
  })
```

@@

```{r plot_gains_and_losses_by_id, fig.width = 12}
# mean proportion loss / gain per chromosome per celltype per id
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::mutate(cell_group = get(params$cell_group_name)) %>%
  dplyr::group_by(sample, cell_group) %>%
  dplyr::summarise(
    dplyr::across(dplyr::starts_with('proportion_'), ~ mean(.x, na.rm = T))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(sample_x_cg = paste0(sample, '_', cell_group)) %>%
  tibble::column_to_rownames('sample_x_cg')

# split gain and loss
c('dupli', 'loss') %>% 
  purrr::map(function(direction) {
    p_mat <-
      p_dat[, grepl(paste0('proportion_', direction), colnames(p_dat))]
    p_annot <- 
      p_dat[, c('sample', 'cell_group')] %>%
      as.data.frame()
    pheatmap::pheatmap(
      p_mat, 
      show_rownames = F, 
      annotation_row = p_annot,
      main = paste0('direction: ', direction)) %>%
      print()
  })
```

@@

```{r chr7_prop_dist}
# prep data
p_dat <-
  sce@colData %>%
  tibble::as_tibble() %>%
  dplyr::group_by(sample) %>%
  dplyr::mutate(cell_group = get(params$cell_group_name),
                n_cells = dplyr::n())

# define functions
plot_prop_chr7_gain <- function(p_dat, title) {
  plot_func <- function(p_dat, position_arg) {
    p_dat %>%
      ggplot() +
      geom_bar(aes(x = cell_group, 
                   fill = proportion_dupli_chr7, 
                   group = proportion_dupli_chr7),
               position = position_arg) +
      scale_x_discrete(guide = guide_axis(angle = -45)) +
      theme_classic() +
      scale_fill_viridis_b(n.breaks = 9, limits = c(0, 1)) 
  }

@@

We will define chr7+ cells as those with >80% of genes affected by duplication and chr7 WT cells as those with <20% of genes affected. We will pseudobulk by patient, cell group, and chr7 ploidy to produce a reduced object.

```{r chr7_gain, class.source = 'fold-show'}
sce$chr7_gain <-
  ifelse(sce$proportion_dupli_chr7 > 0.8, 1, 
         ifelse(sce$proportion_dupli_chr7 < 0.2, 0, 
                NA))
```

```{r chr7_gain_prop, result = 'asis'}
dat <-
  tibble::tibble(chr7_gain_value = sce$chr7_gain,
               cell_group = sce$cell_group) %>%
  dplyr::mutate(
    chr7_gain_status = dplyr::case_when(
      is.na(chr7_gain_value) ~ 'unknown/intermediate',
      chr7_gain_value == 0 ~ 'WT',
      chr7_gain_value == 1 ~ 'gain')
  )

cat('Counts and proportions of chr7 gain, all cells')
dat %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()

cat('Counts and proportions of chr7 gain, by cell group')
dat %>%
  dplyr::group_by(cell_group) %>%
  dplyr::mutate(n_cells = dplyr::n()) %>%
  dplyr::group_by(cell_group, chr7_gain_status, n_cells) %>%
  dplyr::summarise(n = dplyr::n(),
                   `%` = 100 * dplyr::n() / unique(n_cells)) %>%
  knitr::kable()
```

```{r pseudobulk_chr7_gain, class.source = 'fold-show'}
reduced_sce <- xfun::cache_rds({

  # pseudobulk the sce object by patient, cell group and chr7 ploidy
  reduced_sce <-
    glmGamPoi::pseudobulk(
      sce[, !is.na(sce$sample) & 
            !is.na(sce$chr7_gain) &
            !is.na(sce$cell_group)], 
      group_by = vars(sample, cell_group, chr7_gain)
    )

@@

We will identify which genes in stromal cells are changed most by chr7 gain. We do this by fitting one GLM model for each gene, with the chr7 status, cell group and patient as factors.

```{r deseq, message = T}
# fit one GLM model for each gene
fit <- xfun::cache_rds({
  fit <-
    glmGamPoi::glm_gp(
      reduced_sce,
      design = ~ chr7_gain + sample + cell_group,
      verbose = T
    )
  fit
}, file = 'fit.rds', rerun = params$rerun)

@@

  # test differences
  de_res <-
    glmGamPoi::test_de(
      fit, 
      contrast = 
        cond(chr7_gain = 1) -
        cond(chr7_gain = 0)
    )

@@

```{r plot_de}
# plot
de_res %>%
  ggplot(aes(x = lfc, y = -log10(adj_pval), colour = direction, 
             size = chr == 'chr7', shape = chr == 'chr7')) +
  geom_point() +
  geom_hline(yintercept = -log10(min_p), linetype = 'dashed') +
  geom_vline(xintercept = c(log2(1.5), log2(2)), linetype = 'dashed', colour = 'purple') +
  geom_vline(xintercept = c(-1.5, 1.5), linetype = 'dashed') +
  ggrepel::geom_text_repel(
    data = 
      de_res %>% 
      dplyr::filter(!is.na(direction)) %>%
      dplyr::group_by(direction) %>%
      dplyr::filter(rank(adj_pval) <= 10 | rank(lfc, ties.method = 'min') <= 10),
    aes(label = name, ),
    min.segment.length = 0, nudge_x = 0.7, show.legend = F) +
  theme_classic() +
  scale_colour_manual(values = c('upregulated' = 'darkred',
                                 'downregulated' = 'darkblue'), 
                      na.value = 'darkgrey') +
  scale_size_manual(values = c('TRUE' = 3, 'FALSE' = 2), na.value = 2) +
  scale_shape_manual(values = c('TRUE' = 55, 'FALSE' = 16), na.value = 16)
```
